Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-29 sections/tensors.tex
more fixing

\subsection{The 8 Key Level Structures}
    \begin{wrapfigure}{R}{.35\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{wrapfigure}
    The main benefits of specializing to structure come from the following properties of the data:
    \begin{enumerate}
        \item[Sparsity] Sparse data is data that is mostly zero, or some other
        fill value. When we specialize on this data, we can use annihilation ($x
        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
        redundant work.

        \item[Blocks] Blocked data is a subset of sparse data where the nonzeros
        are clustered and occur adjacent to one another. This provides us with
        two opportunities: We can avoid storing the locations of the nonzeros
        individually, and we can use more efficient randomly accessible
        iterators within the block. \cite{im_optimizing_2001, vuduc_performance_2002, ahrens_looplets_2023}.

        \item[Runs] Runs of repeated values may occur in dense or sparse code,
        cutting down on storage and allowing us to use integration rules such as 
        \mintinline{julia}{for i = 1:n; s += x end} $\rightarrow$
        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops \cite{donenfeld_unified_2022,ahrens_looplets_2023}.

        \item[Singletons] When we have only one non-fill region in sparse data,
        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
    \end{enumerate}

    In Finch, we have identified 8 key level structures that can represent all
    of the relevant combinations of these properties, summarized in Table
    \ref{tab:TypesOfStructure}. We examine each structure in turn, describing
    some of the key properties and potential use cases of each. In this sense,
    the structures we consider are exhaustive. We can represent a wide variety
    of hierarchical tensor structures by combining these level structures in a
    tree, as shown in Figure~\ref{fig:structuraldiversity}.

@@

@@

\begin{figure}
    \raggedright
\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
of size $dims$ and an initial value of $init$. Requires the level to be in
read-only mode.
\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
level for reading. Requires the level to be in update-only mode.
\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
to be in read-only mode.
\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
returns a looplet nest over the values in the read-only fiber.  When $mode =
\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
fiber. Often, skipping over mutable locations allows the level to know which
locations must be stored. Additionally, a dirty bit may be used to communicate
whether the mutable subfiber has been written to, which allows the parent fiber
to know whether the subfiber must be stored explicitly.
\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
only ever called on unassembled positions, but some levels (such as dense levels
or bytemaps) may support reassembly.
\caption{The five functions that define a level.}
\end{figure}

\subsection{Level Formats}
Having described the 8 basic level structures and the functions that define a
level format, we can describe the concrete level formats supported by Finch,
some of the challenges in their implementation, and how we overcome them. Note
that these formats are not meant to be exhaustive; we will later give
recommendations for additional level formats to be implemented and explain how
users can also add their own custom formats.

A few challenges arise when implementing these levels. First, it is difficult to merge duplicate runs.
%
Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
%
Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
%
Instead, we merge the duplicates during the freeze phase.
%
We $freeze$ the sublevel, $declare$ a separate sublevel as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer.

Second, we must be able to handle the case where a sublevel is not stored because a parent level is sparse.
%
Therefore, all of Finch's sparse formats use a dirty bit during writing to determine whether the sublevel has been modified from it's default fill value and thus, whether it needs to be stored.

@@

@@

    It also stores a duplicate sublevel, $buf$, for deduplication. \\

@@

    It also stores a duplicate sublevel, $buf$, for deduplication. \\

@@

    \caption{Description of Data Storage Formats}

@@

\begin{enumerate}
\item[SparseHash]
    The sparse hash format uses a hash table to store the locations of nonzeros,
    and sorts the unique indices for iteration during the freeze phase. This
    allows for efficient random access, but not incremental construction, as the
    freeze phase runs in time proportional to the number of nonzeros in the
    entire level.
    \begin{enumerate}
        \item[$idx$] A vector of nonzero indices in the level
        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
        \item[$tbl$] A hash table used for construction of the level.
    \end{enumerate}
\item[SparseBytemap]
    The SparseBytemap format uses a bytemap to store which locations have been
    written to. Unlike the SparseHash format, the bytemap assembles the entire
    space of possible subfibers. This accelerates random access in the format,
    but requires a high memory overhead. Because we don't want to reallocate all
    of the memory in each iteration, the declaration of this format instead
    re-assembles only the dirty locations in the tensor. This format is
    analogous to the default workspace format used by TACO.
    \begin{enumerate}
        \item[$idx$] A vector of nonzero indices in the level
        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
        \item[$tbl$] A dense array of booleans used to mark dirty locations in the tensor.
    \end{enumerate}
\end{enumerate}

>>>>

@@

\begin{figure}[htbp]
    \scriptsize
    \raggedright
\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
of size $dims$ and an initial value of $init$. Requires the level to be in
read-only mode.
\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
level for reading. Requires the level to be in update-only mode.
\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
to be in read-only mode.
\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
returns a looplet nest over the values in the read-only fiber.  When $mode =
\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
fiber. Often, skipping over mutable locations allows the level to know which
locations must be stored. Additionally, a dirty bit may be used to communicate
whether the mutable subfiber has been written to, which allows the parent fiber
to know whether the subfiber must be stored explicitly.
\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
only ever called on unassembled positions, but some levels (such as dense levels
or bytemaps) may support reassembly.
\caption{The five functions that define a level.}
\end{figure}

@@

\subsection{The 8 Key Level Structures}

    The main benefits of specializing to structure come from the following properties of the data:
    \begin{itemize}
        \item \textbf{Sparsity} Sparse data is data that is mostly zero, or some other
        fill value. When we specialize on this data, we can use annihilation ($x
        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
        redundant work.

        \item \textbf{Blocks} Blocked data is a subset of sparse data where the nonzeros
        are clustered and occur adjacent to one another. This provides us with
        two opportunities: We can avoid storing the locations of the nonzeros
        individually, and we can use more efficient randomly accessible
        iterators within the block. \cite{im_optimizing_2001, vuduc_performance_2002, ahrens_looplets_2023}.

        \item \textbf{Runs} Runs of repeated values may occur in dense or sparse code,
        cutting down on storage and allowing us to use integration rules such as 
        \mintinline{julia}{for i = 1:n; s += x end} $\rightarrow$
        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops \cite{donenfeld_unified_2022,ahrens_looplets_2023}.

        \item \textbf{Singletons} When we have only one non-fill region in sparse data,
        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
    \end{itemize}

    \begin{wrapfigure}{r}{.35\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{wrapfigure}

    We consider all combinations of these structural properties in Table
    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
    the the 8 resulting situations. While it is impossible to write code which
    precisely addresses every possible structure, the combination of our level
    formats can express a wide variety of hierarchical structures to a
    sufficient granularity that we can generate efficient code for them.  For
    example, though banded arrays are a superset of ragged arrays, representing
    a ragged array using a banded format does not add much overhead.  The
    structures we consider are exhaustive in the sense that they address all 
    combinations of sparsity, blocks, runs, and singletons in each level. We can
    represent a wide variety of hierarchical tensor structures by combining
    these level structures in a tree, as shown in
    Figure~\ref{fig:structuraldiversity}.

    Having described the 8 basic level structures and the functions that define a
level format, we can describe the concrete level formats supported by Finch,
some of the challenges in their implementation, and how we overcome them. Note
that these formats are not meant to be exhaustive; we will later give
recommendations for additional level formats to be implemented and explain how
users can also add their own custom formats.

A few challenges arise when implementing these levels. First,

Second,

@@

    \scriptsize

@@

    A challenge arises for this level: it is difficult to merge duplicate runs.
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    Instead, we merge the duplicates during the freeze phase.
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\

@@

    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\

@@

    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}

@@
